\texttt{disttree}: Distributional Trees and Forests

### 2017-09-10 pkg/disttree/inst/draft/draft.Rnw
small bugfixes and additional parameter 'censpar'
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@1574 255f4249-2263-4417-b1ea-8498779e27a7

### plotting function for one data set
plot1 <- function(learndata, 
                  dt=NULL, df=NULL, g=NULL, gb=NULL, b=NULL, rf=NULL, cf=NULL, 
                  mu=TRUE, sigma=TRUE) 
{
  plotdata <- cbind(learndata[,c("y","x1","mu","sigma")],
                    df$fitted.par,
                    dt$fitted.par,
                    #coef(dt)[paste(dt$fitted[,1]),],
                    g$mu.fv, g$sigma.fv)

  colnames(plotdata) <- c("y","x1","true.mu","true.sigma", 
                          "fitted.mu.df","fitted.sigma.df",
                          "fitted.mu.dt","fitted.sigma.dt",
                          "fitted.mu.g","fitted.sigma.g")
  sp <- plotdata[order(plotdata["x1"]),]

  if(!(any(dt.par, df.par, g.par, all.par, all.var))){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "True parameters", cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
  col = transpgrey, border = "transparent")
    #lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", lty = 1, col = 'grey')
    #lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", lty = 1, col = 'grey')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$')), 
    #       col = c('black', transpgrey), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # disttree
  if(dt.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "disttree", col.main = pal["tree"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2.5)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
  col = pallight["tree"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = 'greenyellow')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
    #       col = c('black','grey', pal["tree"], pallight["tree"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # distforest
  if(df.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", main = "distforest", col.main = pal["forest"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2.5)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
  col = pallight["forest"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = 'orange')
    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = 'orange')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')), 
    #       col = c('black','grey',pal["forest"], pallight["forest"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # gamlss
  if(g.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey",  main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, xaxt="n", yaxt="n", xlab = "", ylab = "")
    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'grey')
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2.5)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
  col = pallight["gamlss"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = 'lightblue')
    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = 'lightblue')
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\mu \\pm \\sigma$'), 
    #                    TeX('$\\hat{\\mu}$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$')),  
    #       col = c('black','grey',pal["gamlss"], pallight["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

  # disttree vs. distforest vs. gamlss
  if(all.par){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)

    lines(x = sp$x, y = sp$true.mu, type = "l", col = 'black')
    lines(x = sp$x, y = sp$fitted.mu.dt, type = "l", col = pal["tree"], lwd = 2)
    lines(x = sp$x, y = sp$fitted.mu.df, type = "l", col = pal["forest"], lwd = 2)
    lines(x = sp$x, y = sp$fitted.mu.g, type = "l", col = pal["gamlss"], lwd = 2)
    #legend('topleft', c(TeX('$\\mu$'), TeX('$\\hat{\\mu}$ disttree'),
    #                    TeX('$\\hat{\\mu}$ distforest'), TeX('$\\hat{\\mu}$ gamlss')), 
    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
    legend(x = -0.5, y = 13.95, expression(mu), bty = "n")
  }

  # variance: disttree vs. distforest vs. gamlss
  if(all.var){
    par(mar=c(5.1,4.1,4.1,3.1))
    plot(y = sp$y, x = sp$x, type = "p", col="grey", xaxt="n", yaxt="n", xlab = "", ylab = "")
    title(main = "disttree", col.main = pal["tree"], cex.main = 1.2, font.main = 2, adj = 0.19)
    title(main = "distforest", col.main = pal["forest"], cex.main = 1.2, font.main = 2, adj = 0.5)
    title(main = "gamlss", col.main = pal["gamlss"], cex.main = 1.2, font.main = 2, adj = 0.8)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.35)
    title(main = "vs.", col.main = "black", cex.main = 1.2, font.main = 2, adj = 0.65)

    #polygon(c(sp$x, rev(sp$x)), c(sp$true.mu + sp$true.sigma, rev(sp$true.mu - sp$true.sigma)),
    #col = transpgrey, border = "transparent")
    lines(x = sp$x, y = sp$true.mu + sp$true.sigma, type = "l", col = 'black')
    lines(x = sp$x, y = sp$true.mu - sp$true.sigma, type = "l", col = 'black')
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.dt + sp$fitted.sigma.dt, rev(sp$fitted.mu.dt - sp$fitted.sigma.dt)),
  col = pallight["tree"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.dt + sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
    #lines(x = sp$x, y = sp$fitted.mu.dt - sp$fitted.sigma.dt, type = "l", col = pal["tree"], lwd = 2)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.df + sp$fitted.sigma.df, rev(sp$fitted.mu.df - sp$fitted.sigma.df)),
  col = pallight["forest"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.df + sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
    #lines(x = sp$x, y = sp$fitted.mu.df - sp$fitted.sigma.df, type = "l", col = pal["forest"], lwd = 2)
    polygon(c(sp$x, rev(sp$x)), c(sp$fitted.mu.g + sp$fitted.sigma.g, rev(sp$fitted.mu.g - sp$fitted.sigma.g)),
  col = pallight["gamlss"], border = "transparent")
    #lines(x = sp$x, y = sp$fitted.mu.g + sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
    #lines(x = sp$x, y = sp$fitted.mu.g - sp$fitted.sigma.g, type = "l", col = pal["gamlss"], lwd = 2)
    #legend('topleft', c(TeX('$\\mu \\pm \\sigma$'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ disttree'),
    #                    TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ distforest'), TeX('$\\hat{\\mu} \\pm \\hat{\\sigma}$ gamlss')), 
    #       col = c('black',pal["tree"],pal["forest"],pal["gamlss"]), lty = 1, bty = "n", lwd = 2.5)
  legend(x = -0.5, y = 14, expression(mu %+-% sigma), bty = "n")
  }

@@

The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and two covariates $X_1$ and $X_2$.\\
The covariates are generated based on the following distribution assumptions:
\begin{center}
\begin{tabular}{ l  c  l }
$X_1$ & $\sim$ & \textit{Unif}([-0.4,1]) \\ 
$X_2$ & $\sim$ & \textit{Unif}([-10,10])
\end{tabular}
\end{center}
The distribution of the response variable is a normal distribution where the corresponding distribution parameters are defined by parameter functions depending on the covariates. 
\vspace{-1cm}
\begin{center}
$$ Y \sim \mathcal{N} (f(x_1, x_2, \kappa))$$
with
$$f(x_1, x_2, \kappa) = \left(\mu(x_1,x_2,\kappa) , \sigma(x_1,x_2,\kappa)\right)$$ 
\end{center}

@@

Including non-additive effects creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated. Here a non-additive effects is created by separating the generated data into two subgroups depending on the value of $X_2$. In particular, within the function $\mu$ each observation is assigned with a probability of being in subgroup 1. This probability is specified by a smooth function and used to generate a binary variable indicating in which subgroup the observation is. Depending on the subgroup the actual smooth parameter function is defined.
\vspace{-1cm}
\begin{center}
$$\mu(x_1,x_2,\kappa)  = 10 \cdot \Big(1-s(x_1,x_2,\kappa)+ \big(2\cdot s(x_1,x_2,\kappa) - 1\big) \cdot \text{\code{rbinom}}\left(n,1,P_1\right)\Big)
$$
with the distribution function of the logistic distribution as a smooth function
$$
s(x_2,\kappa) = \text{\code{plogis}}\left(\kappa^2 \cdot \frac{x_2 - 3}{10}\right)
$$
and the probability of being in subgroup 1
$$
P_1(x_1,\kappa) = e^{-(10\cdot x_1 - 5)^{2 \cdot \kappa}}.
$$
\end{center}
In this scenerio the scale parameter $\sigma$ depends on the location parameter $\mu$ in the following way:
\vspace{-1cm}
\begin{center}
$$\sigma(x_1,x_2,\kappa)  = 2 + \frac{2 \cdot \mu(x_1,x_2,\kappa)}{5}$$
\end{center}
The  methods are compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models.\\
Generally, a smooth parametric model such as GAMLSS should be the most successful method if the parameter function is smooth with no rapid changes. However, if there are any jumps or non-additive effects or interactions, the trees should be more effective in detecting these steps and therefore also lead to a better fitting forest. But this situation might also appear for smooth functions which contain sections with very steep slopes. Therefore, it is now of great interest, whether the concept of GAMLSS or of trees and forests covers a broader range of functions where it is superior to the other one.\\
Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups. This is the point where the tree structure provides a better solution. However, a single tree can not keep up with the performance of \code{gamlss} when it comes to modeling the smooth functions within the subgroups. Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation. For $\kappa=1$ the performance of all three methods is similar. However, as $\kappa$ increases the RMSE of gamlss increases while the RMSE of trees and forests stays at almost the same level or even decreases sligthly as soon as $\kappa$ is greater or equal 2. The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become. Especially regarding the scale parameter the trees are clearly advantageous over the forests.

>>>>

@@

The different methods are tested on generated data sets consisting of the values of a response variable $Y$ and one or both of the two covariates $X_1$ and $X_2$.\\
The covariates are generated based on the following distribution assumptions:
\begin{center}
\begin{tabular}{ l  c  l }
$X_1$ & $\sim$ & \textit{Unif}([-0.4,1]) \\ 
$X_2$ & $\sim$ & \textit{Unif}([-10,10])
\end{tabular}
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 1 covariate
In a first example the distribution of the response variable is a left censored normal distribution with censoring point 0 where the corresponding distribution parameters are defined by a parameter function $f$ depending on one covariate only.

\begin{center}
$$ Y \sim \text{c}\mathcal{N} (f(x_1, \kappa))$$
with
$$f(x_1, \kappa) = \left(\mu(x_1,\kappa) , \sigma(x_2,\kappa)\right)$$ 
\end{center}

@@

\vspace{-1cm}
\begin{center}
$$
\mu(x_1,\kappa) = 
\begin{cases}
    4 + 8 \cdot \exp{(-(4 \cdot x_1-0.7)^{(2\cdot \kappa)})},& \text{if } x_1 \geq 0\\
    0,              & \text{if } x_1 < 0
\end{cases}
$$

$$
\sigma(x_1,\kappa) = 
\begin{cases}
    1+3 \cdot \|x_1\|,& \text{if } x_1 \geq 0\\
    0.00001,              & \text{if } x_1 < 0
\end{cases}
$$
\end{center}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 2 covariates
In a second example we consider a normaly distributed response variable with the following parameter function depending on two covariates.
%\vspace{-1cm}
\begin{center}
$$ Y \sim \mathcal{N} (f(x_1, x_2, \kappa))$$
with
$$f(x_1, x_2, \kappa) = \left(\mu(x_1,x_2,\kappa) , \sigma(x_1,x_2,\kappa)\right)$$ 
\end{center}

Including non-additive effects and interactions creates a situation in which the main advantages of distributional trees and in particular distributional forests can be investigated. Here a non-additive effects is created by separating the generated data into two subgroups depending on the value of $X_2$. In particular, within the function $\mu$ each observation is assigned with a probability of being in subgroup 1. This probability is specified by a smooth function and used to generate a binary variable indicating in which subgroup the observation is. Depending on the subgroup the actual smooth parameter function is defined.
\vspace{-1cm}
\begin{center}
$$\mu(x_1,x_2,\kappa)  = 10 \cdot \Big(1-s(x_1,x_2,\kappa)+ \big(2\cdot s(x_1,x_2,\kappa) - 1\big) \cdot \text{\code{rbinom}}\left(n,1,P_1\right)\Big)
$$
with the distribution function of the logistic distribution as a smooth function
$$
s(x_2,\kappa) = \text{\code{plogis}}\left(\kappa^2 \cdot \frac{x_2 - 3}{10}\right)
$$
and the probability of being in subgroup 1
$$
P_1(x_1,\kappa) = e^{-(10\cdot x_1 - 5)^{2 \cdot \kappa}}.
$$
\end{center}
In this scenerio the scale parameter $\sigma$ depends on the location parameter $\mu$ in the following way:
\vspace{-1cm}
\begin{center}
$$\sigma(x_1,x_2,\kappa)  = 2 + \frac{2 \cdot \mu(x_1,x_2,\kappa)}{5}$$
\end{center}
The  methods are compared by the out-of-sample RMSE for the location parameter $\mu$ and the shape parameter $\sigma$ as well as by the out-of-sample log-likelihood of the fitted models.\\
\\
Generally, a smooth parametric model such as GAMLSS should be the most successful method if the parameter function is smooth with no rapid changes. However, if there are any jumps or non-additive effects or interactions, the trees should be more effective in detecting these steps and therefore also lead to a better fitting forest. But this situation might also appear for smooth functions which contain sections with very steep slopes. Therefore, it is now of great interest, whether the concept of GAMLSS or of trees and forests covers a broader range of functions where it is superior to the other one.\\
\\
Even though \code{gamlss} can easily deal with each of the smooth functions it struggles with the separation in subgroups. This is the point where the tree structure provides a better solution. However, a single tree can not keep up with the performance of \code{gamlss} when it comes to modeling the smooth functions within the subgroups. Offering a good compromise the forest takes advantages of the tree structure to find subgroups but can also model smooth effects if enough trees are included.\\
However, once the increasing value of $\kappa$ leads to very steep sections within the parameter functions of each of the subgroups the tree turns out to be the most successful method in this specific situation.\\
\\
Looking at the resulting values of the RMSE it can be observed that trees and forests cleary outperform gamlss in this situation. For $\kappa=1$ the performance of all three methods is similar. However, as $\kappa$ increases the RMSE of gamlss increases while the RMSE of trees and forests stays at almost the same level or even decreases sligthly as soon as $\kappa$ is greater or equal 2. The same conclusion can be drawn from looking at the log-likelihood where the resulting values of gamlss decrease rapidly while the values of trees and forests only decrease a little at the beginning but then increase.\\
This is the result that was expected as trees and forests can deal with rapid changes of the distribution functions more easily.\\
For low values of $\kappa$ the forests are a little bit more efficient than the trees, but the higher $\kappa$ and therefore also the steeper the parameter function is the better the trees become. Especially regarding the scale parameter the trees are clearly advantageous over the forests.
